---
title: "Description of Workflow"
author: "Stephen Pederson"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

```{r packages}
library(tidyverse)
library(DiagrammeR)
library(pander)
library(yaml)
```

```{r config}
config <- here::here("config/config.yml") %>%
  read_yaml()
```


# Pre-Processing Workflow

```{r nodeFuns}
rm_dot_node <- function(x, node = "all"){
  nd <- x[grepl(node, x)]
  id <- str_replace_all(nd, "\\t([0-9]*).label.+", "\\1")
  pat <- paste(c("(", node, "| ", id, "$)"), collapse = "")
  x[!grepl(pat, x)]
}
add_input_node <- function(x, node = "raw_data", ignore = "(get|build|make)", col = "0.00 0.00 0.00", style = "rectangle"){
  # Find the nodes with no input
  all_id <- x %>% 
    str_subset("label") %>% 
    str_replace_all("\\t([0-9]*).label.+", "\\1")
  has_parent <- x %>% 
    str_subset("[0-9]$") %>% 
    str_extract("[0-9]*$")
  to_ignore <- x %>% 
    str_subset(ignore) %>% 
    str_replace_all("\\t([0-9]*).label.+", "\\1")
  no_parent <- setdiff(all_id, c(has_parent, to_ignore))
  # Find the position to insert the node
  new_id <- as.character(max(as.numeric(all_id)) +1)
  new_text <- c(
    "\t", new_id, 
    "[label = \"", node, 
    "\", color = \"", col,
    "\", style=\"", style,
    "\"];"
    ) %>%
    paste(collapse = "")
  last_label <- x %>%
    str_detect("label") %>%
    which() %>%
    max()
  ## Add the node
  x <- c(
    x[seq_len(last_label)], new_text, x[seq(last_label + 1, length(x))]
  )
  new_edges <- paste0("\t", paste(new_id, no_parent, sep = " -> "))
  c(
    x[seq(1, length(x) - 1)],
    new_edges,
    x[length(x)]
  )
}
```

## Workflow Outline


This is an analysis of RNA-Seq data with the data processing workflow derived from https://github.com/steveped/snakemake_rnaseq_se.
The pre-processing workflow as implemented in this analysis is as follows:

```{r plot_workflow, fig.cap = "*Summary of workflow. Node names indicate snakemake rules*"}
here::here("rules/rulegraph.dot") %>%
  readLines() %>%
  rm_dot_node() %>%
  add_input_node(node = "RAW DATA") %>%
  grViz()
```

## Parameter Description

Parameters set for the complete workflow are:

```{r printYAML, echo=FALSE, results='asis'}
pander(config)
```

# Data Description

```{r samples}
samples <- config$samples %>%
  here::here() %>%
  read_tsv() %>%
  mutate_if(
    function(x){length(unique(x)) < length(x)},
    as.factor
  )
n_fact <- sum(vapply(samples, is.factor, logical(1)))
```

A total of `r nrow(samples)` individual fastq files were used as input for the workflow.
Given the sample sheet, `r ifelse(n_fact, "potential predictor variables were identified as:", "no potential predictor variables were identified.")`

```{r showFactors, echo=FALSE}
if (n_fact){
  samples %>%
    select_if(is.factor) %>%
    lapply(levels) %>%
    pander()
}
```
