---
title: "Differential Gene Expression"
author: "Stephen Pederson"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.height = 7,
  fig.width = 10
)
```

```{r packages}
library(tidyverse)
library(yaml)
library(scales)
library(pander)
library(glue)
library(edgeR)
library(AnnotationHub)
library(ensembldb)
```

```{r options}
panderOptions("table.split.table", Inf)
panderOptions("big.mark", ",")
theme_set(theme_bw())
```

```{r config}
config <- here::here("config/config.yml") %>%
  read_yaml()
suffix <- paste0(config$tag)
sp <- config$ref$species %>%
  str_replace("(^[a-z])[a-z]*_([a-z]+)", "\\1\\2") %>%
  str_to_title()
```

```{r samples}
samples <- config$samples %>%
  here::here() %>%
  read_tsv() %>%
  mutate(
    Filename = paste0(sample, suffix)
  ) %>%
  mutate_if(
    function(x){length(unique(x)) < length(x)},
    as.factor
  )
```

```{r check_config}
config$analysis <- config$analysis %>%
  lapply(intersect, y = colnames(samples)) %>%
  .[vapply(., length, integer(1)) > 0]
```

```{r add_columns}
if (length(config$analysis)) {
  samples <- samples %>%
    unite(
      col = group, 
      any_of(as.character(unlist(config$analysis))), 
      sep = "_", remove = FALSE
    ) 
} else {
  samples$group <- samples$Filename
}
```

```{r group_cols}
group_cols <- hcl.colors(
  n = length(unique(samples$group)), 
  palette = "Zissou 1"
  ) %>%
  setNames(unique(samples$group))
```

# Setup

## Annotations

```{r ah}
ah <- AnnotationHub() %>%
  subset(rdataclass == "EnsDb") %>%
  subset(str_detect(description, as.character(config$ref$release))) %>%
  subset(genome == config$ref$build)
stopifnot(length(ah) == 1)
```

```{r ensDb}
ensDb <- ah[[1]]
genesGR <- read_rds(here::here("output/genesGR.rds"))
```

Gene annotations were again loaded from Ensembl Release `r ensemblVersion(ensDb)`.
The [previously defined](qc_aligned.html#Annotation_Setup) `GenomicRanges` object containing GC content and Gene Length was also loaded,
containing information for `r comma(length(genesGR))` genes.

## Count Data

```{r counts}
counts <- here::here("data/aligned/counts/counts.out") %>%
  read_tsv(comment = "#") %>%
  dplyr::select(Geneid, ends_with("bam")) %>%
  rename_at(vars(ends_with("bam")), dirname) %>%
  rename_all(basename) %>%
  column_to_rownames("Geneid")
```

```{r fullDGE}
dge <- DGEList(
  counts = counts,
  samples = column_to_rownames(samples, "Filename")[colnames(counts),],
  genes = mcols(genesGR[rownames(counts)])[, c("gene_id", "gene_name", "gene_biotype", "entrezid", "longest_tx", "ave_tx_len", "gc_content")]
) %>%
  calcNormFactors()
```

Prior to filtering for undetectable genes, counts were loaded as a `DGEList`, incorporating both sample and gene metadata.
